---
title: "Classification"
author: "Jaime Davila"
date: "2/20/2022"
output:
  pdf_document: default
  html_document: default
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
library(tidyverse)
library(caret)
library(dslabs)
```


## The MNIST dataset

The MNIST database (Modified National Institute of Standards and Technology database) is a large collection of handwritten digits used by the Machine learning community. The `dslabs` packages has a handy function called `read_mnist` that allows to load this dataset as follows:

```{r}
mnist <- read_mnist("~/Mscs 341 S22/Class/Data")
```

The first thing to notice about this dataset is its structure which we can find using the function `str`

```{r}
str(mnist)
```

As we can wee `mnist` has a training and testing set. The training dataset has 60,000 elements represented as a matrix of $6000 \times 784$ (every image is a vector of 784, representing a $28 \times 28$ image). It also has the labels corresponding to each of the images represented as integers. Finally the testing dataset has 10,000 elements represented in a similar way. Before we interact with this dataset, let's define a handy function that will allows us to plot any digit:

```{r}
plotImage <- function(dat,size=28){
  imag <- matrix(dat,nrow=size)[,28:1]# 28:1 is number of columns to keep
  image(imag,col=grey.colors(256), xlab = "", ylab="") 
}

mnist$train$images[10,]#a single row is all the data. function above rewrites it into a matrix of 28x28
plotImage(mnist$train$images[10,])#
```

So now let's explore a couple of elements from our training and testing datasets

```{r}
# We plot the 10th element from our training dataset which is a 4.
plotImage(mnist$train$images[10,])#the input represents a column of data from original
mnist$train$labels[10]

# We plot the 102th element from our training dataset which is a 7.
plotImage(mnist$train$images[102,])
mnist$train$labels[102]

# We plot the 212th element from our testing dataset which is a 5.
plotImage(mnist$test$images[212,])
mnist$test$labels[212]
```

## Is it a 2 or a 7?

Our original problems has 784 predictors and a response variable with 10 different levels corresponding to each digit. We will start by simplifying our problem/dataset to recognize whether a digit is a `2` or a `7` and we will be using only two predictors, namely:

* `x_1` will be the proportion of dark pixels in the upper left quadrant.
* `x_2` will be the proportion of dark pixels in the lower right quadrant.

Conveniently for us `dslabs` contains a random sample of 1000 digits (800 in training and 200 in testing). Let's load the dataset and plot it

```{r}
data("mnist_27")
str(mnist_27)
train.tbl <- tibble(mnist_27$train)
train.tbl <-  train.tbl %>%
  mutate(n=row_number())

test.tbl <- tibble(mnist_27$test)
test.tbl <-  test.tbl %>%
  mutate(n=row_number())

ggplot(train.tbl, aes(x=x_1, y=x_2, color=y))+
  geom_point()
```

1. Pick the elements with the largest and smallest values in  `x_1` and plot their corresponding images. Do a boxplot comparing the value of `x_1` across 2 and 7s. Does `x1` allow you to distinguish in general between a 2 and 7?. Do the same analysis for `x_2`.

*Hint*:Notice that the corresponding index in the `mnist` dataset can be found by using `mnist_27$index_train`

```{r echo=FALSE}
max.index = slice_max(train.tbl,x_1) %>% pull(n)
index.mnist = mnist_27$index_train[max.index]
plotImage(mnist$train$images[index.mnist,])

min.index = slice_min(train.tbl,x_1, with_ties = FALSE) %>% pull(n)
index.mnist = mnist_27$index_train[min.index]
plotImage(mnist$train$images[index.mnist,])

ggplot(train.tbl, aes(y, x_1, fill=y)) +
  geom_boxplot()
```

```{r echo=FALSE}
max.index = slice_max(train.tbl,x_2) %>% pull(n)
index.mnist = mnist_27$index_train[max.index]
plotImage(mnist$train$images[index.mnist,])

min.index = slice_min(train.tbl,x_2, with_ties = FALSE) %>% pull(n)
index.mnist = mnist_27$index_train[min.index]
plotImage(mnist$train$images[index.mnist,])

ggplot(train.tbl, aes(y, x_2, color=y)) +
  geom_boxplot()
```

# Classification and KNN

We can build our first model by using a KNN model. Notice that since we are in the classification setting we will be using the function `knn3`

```{r}
kNear=5
knn.model <- knn3(y~x_1+x_2, data=train.tbl, k=kNear)
```

Notice that we can use the function `predict` on our testing dataset

```{r}
pred.prob <- predict(knn.model, test.tbl)
head(pred.prob)
```

If we are interested in obtaining a class label we can do so by using the parameter `type="class"`

```{r}
pred <- predict(knn.model, test.tbl, type="class")
head(pred)
```

2. Plot the probability that an element is a `2` on the testing dataset

```{r echo=FALSE}
test.tbl <- test.tbl %>%
  mutate(pred.2 = pred.prob[,1],
         pred.7 = pred.prob[,2])

ggplot(test.tbl, aes(y,pred.2, fill=y))+
  geom_boxplot()

ggplot(test.tbl, aes(y,pred.7, fill=y))+
  geom_boxplot()
```

3. Create a function `calc_error (kNear, train, test)` that calculates the misclassification error for KNN using `knear` neighbors

```{r echo=FALSE}
calc_error <- function(kNear, train.tbl, test.tbl) {
  
  knn.model <- knn3(y~x_1+x_2, data=train.tbl, k=kNear)
  pred <- predict(knn.model, test.tbl, type="class") 
  
  mean (pred!=test.tbl$y)
}

calc_error(5, train.tbl, test.tbl)
calc_error(50, train.tbl, test.tbl)
calc_error(200, train.tbl, test.tbl)
```

4. Plot the value of `k` against the misclassification error for $k=1..100$ using the testing dataset. What is the optimal value of `k`?


```{r echo=FALSE}
error.test <- vector(length = 100) 
for(i in 1:100){
  error.test[i] <- calc_error(i, train.tbl, test.tbl)
}
error.tbl <- tibble (k=1:100,
                     error = error.test)

error.tbl %>%
  ggplot(aes(x=k, y=error)) +
  geom_line()

min.k.tbl <- error.tbl %>% 
  slice_min(error, with_ties = FALSE)
min.k.tbl
```

5. (*Optional*) Using the optimal value of `k` identify cases that are missclassified. Is it more common that 2 gets confused for a 7 or the other way around? Plot the values of `x_1` and `x_2` for those.
Plot a couple of digits that are missclassified. Any ideas for features that would allow for the correct classification?

```{r echo=FALSE}
kNear=40
knn.model <- knn3(y~x_1+x_2, data=train.tbl, k=kNear)
pred.label <- predict(knn.model, test.tbl, type="class") 

test.tbl <- test.tbl %>%
  mutate(pred.fct=pred.label)
  
table(test.tbl$y, test.tbl$pred.fct)

test.tbl %>%
  filter(pred.fct!=y) %>%
  ggplot(aes(x_1,x_2,color=y)) +
  geom_point()

#2s that got labeled as 7s
index <- test.tbl %>%
  filter(pred.fct=="7" & y =="2") %>%
  arrange(desc(x_1)) %>%
  pull(n)

(index.mnist = mnist_27$index_test[index])

plotImage(mnist$train$images[index.mnist[1],])
plotImage(mnist$train$images[index.mnist[2],])

#7s that got labeled as 2s

index <- test.tbl %>%
  filter(pred.fct=="2" & y =="7") %>%
  arrange(x_1) %>%
  pull(n)

(index.mnist = mnist_27$index_test[index])

plotImage(mnist$train$images[index.mnist[1],])
plotImage(mnist$train$images[index.mnist[3],])
```


